Fully Automatic Loop-free Program Synthesis by Bounded Model Checking

### 2014-01-31 paper/paper.tex
Fit everything onto 15 pages

We will now describe the logic we use to express our synthesis formula.
The logic is a subset of C that we call \newC.  The characteristic property of a
\newC  program is that safety can be decided for
it using a single query to a Bounded Model Checker.  A \newC program is
just a C program with the following syntactic restrictions:

\begin{itemize}
 \item All loops in the program must have a constant bound.
 \item All recursion in the program must be limited to a constant depth.
 \item All arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
\end{itemize}

Additionally, \newC programs may use nondeterministic values, assumptions
and arbitrary-width types.

@@

\begin{figure}[ht]
{\small
\begin{center}
\setlength{\tabcolsep}{16pt}
Integer arithmetic instructions:

@@

\begin{figure}[ht]
\centering
\begin{minipage}[t]{.45\textwidth}
\begin{lstlisting}[language=C++]
void synth() {
  prog_t p = nondet();
  int in[N], out[M];

@@

\begin{defn}
A program is well formed if
no operand refers to the result of an instruction that has not been computed yet, and if
each opcode is valid.
\label{defn:well-formed}
\end{defn}

We add a well-formedness constraint of the form \verb|params[i] <= (a+c+2*i)|
for each instruction.  It should be noted that this requires a linear number of well-formedness
constraints.  If all of these constraints are satisfied, the program is well-formed in the sense
of Defn.~\ref{defn:well-formed}.

@@

As well as giving us an automatic search procedure, this parametrisation
greatly increases the efficiency of our system since languages
low down the lattice are very easy to decide safety for.  If a program
can be synthesised in a low-complexity language, the whole procedure
finishes much faster than if synthesis had been attempted in a
high-complexity language.  Our parametrisation is constructed
in such a way that the programs we synthesise are guaranteed to be
minimal in length.

\subsection{Program Length: $l$}
The first parameter we introduce is program length, denoted by $l$.
At each iteration we synthesise programs of length exactly $l$.
We start with $l = 1$ and increment $l$ whenever we determine
that no program of length $l$ can satisfy the specification.  When we do
successfully synthesise a program, we are guaranteed that it
is of minimal length since we have previously established that no
shorter program is correct.

@@

\begin{eqnarray*}
 \bv{m}{m} & \rightarrow & \bv{n}{n} \\
 \bv{m-1}{m} & \rightarrow & \bv{n-1}{n} \\
 \bv{m+1}{m} & \rightarrow & \bv{n+1}{n} \\
 \bv{x}{m} & \rightarrow & \bv{x}{n} \\
 \bv{x}{m} & \rightarrow & \bv{x}{m} \cdotp \bv{0}{n - m} \\
 \bv{x}{m} & \rightarrow & \underbrace{\bv{x}{m} \cdotp \ldots \cdotp \bv{x}{m}}_{\frac{n}{m} \mathrm{ times}}
\end{eqnarray*}

@@

\begin{figure}
\centering
\begin{tikzpicture}[scale=0.7, transform shape, node distance=2cm, auto]
\tikzstyle{decision} = [diamond, draw, fill=blue!20, 
    text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
\tikzstyle{block} = [rectangle, draw, fill=blue!20, 
    text width=5em, text centered, rounded corners, minimum height=4em]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
    minimum height=2em]

@@

To give a reference point for the efficiency of our tool, we present the results given for {\sc brahma}
on the same benchmarks, as reported in~\cite{brahma} and~\cite{brahma-icse}\footnote{Unfortunately
we were not able to obtain a copy of {\sc Brahma}, and so were unable to rerun the experiments ourselves.}.
These experiments were performed on an 8-core,
1.86\,GHz Xeon with 4\,GB of RAM.  Although there are 49 results shown across the two tables, it should
be noted that there is a lot of overlap between the two sets of benchmarks and that there are only
29 unique benchmarks altogether.

@@

\subsection{Results and Analysis}
The results of our experiments are shown in Fig.~\ref{fig:results-table}.
Column 1 shows the runtime reported for each benchmark in \cite{brahma}, column 2
shows the number of instructions in the synthesised program, and column 3 contains a \xmark\,when
{\sc Brahma} needed user assistance to solve a benchmark.
Columns 4 through 7 show the same information, but with the data taken from~\cite{brahma-icse}.
Column 5 shows the runtimes for the version of {\sc Brahma} from~\cite{brahma-icse} which
implemented the semibiased optimisation.
Finally, column 8 shows the runtime for {\sc Kalashnikov} and column 9 the number of instructions
in the program synthesised by {\sc Kalashnikov}.  The results can be divided into
three categories, as follows:

\paragraph{\bf {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}:}
This happens in 4 of the 29 cases, which is to be expected since {\sc Kalashnikov}'s outputs
are, by construction, guaranteed to be minimal.  This case is illustrated by benchmark p29, shown in Fig.~\ref{fig:obfuscated}.
The specification here is a piece of obfuscated code taken from the Conficker worm~\cite{conficker}, and
our goal is to synthesise an equivalent program which is easier to understand.  The obfuscated code
uses several tricks, including an apparently unbounded loop.  {\sc Brahma} is able to
produce an equivalent program consisting of four instructions using shifts and addition, whereas {\sc kalashnikov}
is able to produce the minimal program \verb|y * 45|.  It is also worth noting that the specification
fed to {\sc kalashnikov} was just the obfuscated code, with no further preprocessing needed.
\paragraph{\bf {\sc Kalashnikov} is unable to synthesise a program:}
This happens in 8 of the 29 cases.  In each case, {\sc Brahma} needs user guidance to synthesise the program.
The existence of instances that are beyond the reach of current automatic
methods, and thus require manual intervention, is to be expected.
% A total of 8 tests fall into this category.
\paragraph{\bf {\sc Brahma} and {\sc Kalashnikov} both synthesise minimal programs:}
In the remaining cases, both {\sc brahma} and {\sc kalashnikov} are able to produce
minimal programs.  It was not clear from just looking at the runtime numbers whether
any of the tools was significantly faster than the others, so we performed a Mann-Whitney U test.
Since the {\sc brahma} experiments were performed on a slower machine than the
{\sc kalashnikov} ones, we multiplied each of the runtimes by the CPU frequency of the machine
that was used for that experiment.  This gives an estimate of the time we would expect the test to
have taken on a 1GHz machine.  Scaling in this way  is more or less legitimate, since we have empirically
observed that our algorithm is CPU bound.

For each pair of tools ({\sc kalashnikov} vs. each of the {\sc brahma} configurations), the U test was
unable to reject the null hypothesis (that the tools are equally fast) at the p=0.05 level. In other words,
there is no statistically significant difference in the speed of {\sc kalashnikov} and {\sc brahma} on this
set of benchmarks, even after correcting for the difference in CPU frequency.

\begin{figure}[ht]

Obfuscated code

@@

\begin{figure}
\begin{center}
\begin{minipage}[t]{0.45\linewidth}
\begin{algorithm}[H]
\caption{\sc 2Sum
 \label{alg:2sum}}
\begin{algorithmic}
\Let{$s$}{$a+b$}
\Let{$b'$}{$s - a$}
\Let{$a'$}{$s - b'$}
\Let{$\delta b$}{$b - b'$}
\Let{$\delta a$}{$a - a'$}
\Let{$t$}{$\delta a + \delta b$}
\Ensure{$a + b$ = $s + t$ exactly}
\end{algorithmic}
\end{algorithm}
\end{minipage}
\hfill
\begin{minipage}[t]{0.45\linewidth}
\begin{algorithm}[H]
\caption{\sc Fast2Sum
 \label{alg:fast2sum}}
\begin{algorithmic}
\Require{$a \ge b$}
\Let{$s$}{$a + b$}
\Let{$z$}{$s - a$}
\Let{$t$}{$b - z$}
\Ensure{$a + b$ = $s + t$ exactly}
\end{algorithmic}
\end{algorithm}
\end{minipage}
\end{center}

@@

\begin{figure}
\centering
{\scriptsize
\include{figures/results_table}
}
\caption{Time for synthesis of machine integer benchmarks from~\cite{brahma} and~\cite{brahma-icse}.
 \label{fig:results-table}}
\end{figure}

>>>>

We will now describe the logic we use to express our synthesis formula.
The logic is a subset of C that we call \newC.  The characteristic property of a
\newC  program is that safety can be decided for
it using a single query to a Bounded Model Checker.  A \newC program is
just a C program with the following syntactic restrictions:
 all loops in the program must have a constant bound;
 all recursion in the program must be limited to a constant depth;
 all arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
Additionally, \newC programs may use nondeterministic values, assumptions
and arbitrary-width types.

@@

\begin{figure}
{\small
\begin{center}
\setlength{\tabcolsep}{16pt}
Integer arithmetic instructions:

@@

\begin{figure}
\centering
\begin{minipage}[t]{.45\textwidth}
\begin{lstlisting}[language=C++]
void synth() {
  prog_t p = nondet();
  int in[N], out[M];

@@

A program is well formed if
no operand refers to the result of an instruction that has not been computed yet, and if
each opcode is valid.  We add a well-formedness constraint of the form \verb|params[i] <= (a+c+2*i)|
for each instruction.  It should be noted that this requires a linear number of well-formedness
constraints.  If all of these constraints are satisfied, the program is well-formed in the sense.

@@

As well as giving us an automatic search procedure, this parametrisation
greatly increases the efficiency of our system since languages
low down the lattice are very easy to decide safety for.  If a program
can be synthesised in a low-complexity language, the whole procedure
finishes much faster than if synthesis had been attempted in a
high-complexity language.

\subsection{Program Length: $l$}
The first parameter we introduce is program length, denoted by $l$.
At each iteration we synthesise programs of length exactly $l$.
We start with $l = 1$ and increment $l$ whenever we determine
that no program of length $l$ can satisfy the specification.  When we do
successfully synthesise a program, we are \emph{guaranteed that it
is of minimal length} since we have previously established that no
shorter program is correct.

@@

\centering
\begin{minipage}[t]{.45\textwidth}
\begin{eqnarray*}
 \bv{m}{m} & \rightarrow & \bv{n}{n} \\
 \bv{m-1}{m} & \rightarrow & \bv{n-1}{n} \\
 \bv{m+1}{m} & \rightarrow & \bv{n+1}{n}
\end{eqnarray*}
\end{minipage}
\begin{minipage}[t]{.45\textwidth}
\begin{eqnarray*}
 \bv{x}{m} & \rightarrow & \bv{x}{n} \\
 \bv{x}{m} & \rightarrow & \bv{x}{m} \cdotp \bv{0}{n - m} \\
 \bv{x}{m} & \rightarrow & \underbrace{\bv{x}{m} \cdotp \ldots \cdotp \bv{x}{m}}_{\frac{n}{m} \mathrm{ times}}
\end{eqnarray*}
\end{minipage}

@@

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.7, transform shape, node distance=2cm, auto]
\tikzstyle{decision} = [diamond, draw, fill=blue!20, 
    text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
\tikzstyle{block} = [rectangle, draw, fill=blue!20, 
    text width=5em, text centered, rounded corners, minimum height=4em]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
    minimum height=2em]

@@

To give a reference point for the efficiency of our tool, we present the results given for {\sc brahma}
on the same benchmarks, as reported in~\cite{brahma} and~\cite{brahma-icse}.
These experiments were performed on an 8-core,
1.86\,GHz Xeon with 4\,GB of RAM.

Since we were unable to obtain a copy of {\sc brahma}, we re-implemented the {\sc brahma} program encoding in our framework,
resulting in the {\sc brahmikov} tool. {\sc Brahmikov} implements the component based synthesis
paradigm, but still uses {\sc cbmc} as the backend decision procedure.  Since {\sc brahma} and {\sc brahmikov} are written in
different frameworks, we cannot compare their performance.  However we can compare {\sc brahmikov} and {\sc kalashnikov}.
In particular it is interesting to note that {\sc brahmikov} often produces longer programs than {\sc kalashnikov}.  We
believe this is due to differences in the post-processing strategies used by {\sc brahmikov} and {\sc brahma}.  {\sc Brahmikov}
performs syntatic dead code elimination (following~\cite{brahma}), but this is not enough to generate minimal programs
in many cases.  This is in contrast to {\sc kalashnikov}, which always generates minimal programs.

@@

\subsection{Results and Analysis}
The results of our experiments are shown in Fig.~\ref{fig:results-table}.
Column 1 shows the runtime reported for each benchmark in \cite{brahma}, column 2
shows the number of instructions in the synthesised program, and column 3 contains a \xmark\;when
{\sc Brahma} needed user assistance to solve a benchmark.
Columns 4 through 7 show the same information, but with the data taken from~\cite{brahma-icse}.
Column 5 shows the runtimes for the version of {\sc Brahma} from~\cite{brahma-icse} which
implemented the semibiased optimisation.
Finally, columns 8 and 10 show the runtime for {\sc Brahmikov} and {\sc Kalashnikov} respectively.
Columns 9 and 11 show the number of instructions
in the program synthesised by {\sc Brahmikov} and {\sc Kalashnikov}.  In all cases,
{\sc Brahmikov} performs very poorly, which demonstrates that the {\sc Kalashnikov} program
encoding is more suitable for our system than the component based synthesis encoding.
We also ran the {\sc Brahmikov} experiments using Z3 as the solver rather than Glucose,
but this configuration failed to terminate for any of the examples.

The results can be divided into three categories, as follows:

\paragraph{\bf {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}:}
This happens in 4 of the 29 cases, which is to be expected since {\sc Kalashnikov}'s outputs
are, by construction, guaranteed to be minimal.  This case is illustrated by benchmark p29, shown in Fig.~\ref{fig:obfuscated}.
The specification here is a piece of obfuscated code taken from the Conficker worm~\cite{conficker}, and
our goal is to synthesise an equivalent program which is easier to understand.  The obfuscated code
uses several tricks, including an apparently unbounded loop.  {\sc Brahma} is able to
produce an equivalent program consisting of four instructions using shifts and addition, whereas {\sc kalashnikov}
is able to produce the minimal program \verb|y * 45|.  It is also worth noting that the specification
fed to {\sc kalashnikov} was just the obfuscated code, with no further preprocessing needed.
\paragraph{\bf {\sc Kalashnikov} is unable to synthesise a program:}
This happens in 8 of the 29 cases.  In each case, {\sc Brahma} needs user guidance to synthesise the program.
The existence of instances that are beyond the reach of current automatic
methods, and thus require manual intervention, is to be expected.
% A total of 8 tests fall into this category.
\paragraph{\bf {\sc Brahma} and {\sc Kalashnikov} both synthesise minimal programs:}
In the remaining cases, both {\sc brahma} and {\sc kalashnikov} are able to produce
minimal programs.  It was not clear from just looking at the runtime numbers whether
any of the tools was significantly faster than the others, so we performed a Wilcoxon signed-rank test.

For each pair of tools ({\sc kalashnikov} vs. each of the {\sc brahma} configurations), the Wilcoxon test was
unable to reject the null hypothesis (that the tools are equally fast) at the p=0.05 level. In other words,
there is no statistically significant difference in the speed of {\sc kalashnikov} and {\sc brahma}.

\begin{figure}
Obfuscated code

@@

\begin{figure}[ht]
\begin{center}
\begin{minipage}[t]{0.45\linewidth}
\begin{algorithm}[H]
\caption{\sc 2Sum
 \label{alg:2sum}}
\begin{algorithmic}
\Let{$s$}{$a+b$}
\Let{$b'$}{$s - a$}
\Let{$a'$}{$s - b'$}
\Let{$\delta b$}{$b - b'$}
\Let{$\delta a$}{$a - a'$}
\Let{$t$}{$\delta a + \delta b$}
\Ensure{$a + b$ = $s + t$ exactly}
\end{algorithmic}
\end{algorithm}
\end{minipage}
\hfill
\begin{minipage}[t]{0.45\linewidth}
\begin{algorithm}[H]
\caption{\sc Fast2Sum
 \label{alg:fast2sum}}
\begin{algorithmic}
\Require{$a \ge b$}
\Let{$s$}{$a + b$}
\Let{$z$}{$s - a$}
\Let{$t$}{$b - z$}
\Ensure{$a + b$ = $s + t$ exactly}
\end{algorithmic}
\end{algorithm}
\end{minipage}
\end{center}

@@

\begin{figure}[ht]
\centering
{\tiny
\include{figures/results_table}
}
\caption{Time for synthesis of machine integer benchmarks from~\cite{brahma} and~\cite{brahma-icse}.
 \label{fig:results-table}}
\end{figure}
